Implementing simplified INLA into `aghq`

### 2022-11-23 src/dev_sinla/sinla.Rmd
You don't need a new MakeADFun for every z!

The AGHQ/ELGM approximation to the joint posterior of the latent field is the following object
$$
\begin{equation}
\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega(z), (\#eq:pxgiveny)
\end{equation}
$$
where $\theta(z) \in \Theta$ and $\omega(z) \in \mathbb{R}$ are AGHQ nodes and weights generated by "adaption" to $\tilde p_\text{LA}(\theta \, | \, y)$.
In particular, let
$$
\mathcal{Q}(1, k) = \{z_d \in \mathbb{R}: H_k(z_d) = 0\}
$$
be the set of nodes in one-dimension, exactly the zeros of the Hermite polynomials $H_k(z_d)$ for $k \in \mathbb{N}$, and
$$
\omega(z_d) = \frac{k!}{H_{k + 1}(z_d)^2 \times \phi(z_d)}, 
$$
be the associated weights.
Define $\mathcal{Q}(d, k) = \mathcal{Q}(1, k)^d$ to be the set of nodes $z$ in $d$-dimensions, of which there will be $k^d$.
Then these points are adapted via $\theta(z) = \hat \theta + Lz$, where $\hat \theta$ is the mode (of $p_\text{LA}(\theta \, | \, y)$) and $L$ is the lower Cholesky triangle of the curvature (at the mode) such that $LL^\top = H^{-1}$.

@@

We check here that `obj$fn` indeed takes as input $(x_i, \theta)$:

@@

transformed_param <- transform_param(param)
transformed_param$phi_prev_minus_i <- NULL #' Remove the integrated out effects
transformed_param

obj$fn(transformed_param)
```

@@

ks_results %>%
  select(index, D, type) %>%
  pivot_wider(
    names_from = type,
    values_from = D
  ) %>%
  ggplot(aes(x = laplace, y = gaussian)) +
    geom_point(alpha = 0.5) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
    lims(x = c(0, 1), y = c(0, 1)) +
    labs(x = "KS(laplace, tmbstan)", y = "KS(gaussian, tmbstan)") +
    theme_minimal()
```

```{r}
ks_results %>%
  group_by(type) %>%
  summarise(D = mean(D))
```

@@

To start with, let's evaluate this on the log-scale for `i = 1`, `x = x_test_value` and the first value of `z`.
The weight $\omega(z)$ will be `modesandhessians$weights[z]`.
The normalised Laplace approximation $\tilde p_\text{LA}(\theta(z) \, | \, y)$ will be `- h_norm(as.numeric(modesandhessians[z, thetanames]))`.
And to obtain the normalised Gaussian approximation $p_\text{G}(x_i \, | \, \theta(z), y)$ on the log-scale we get the relevant subcomponents of the joint Gaussian then use `ndorm` with `log = TRUE`.

@@

lps <- sapply(gg$nodes, laplace_marginal)
plot_marginal_spline(lps)
```

>>>>

An AGHQ approximation to the joint posterior of the latent field is given by
$$
\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega(z), (\#eq:pxgiveny)
$$
where $\theta(z) \in \Theta$ and $\omega(z) \in \mathbb{R}$ are nodes and weights generated by adaption to $\tilde p_\text{LA}(\theta \, | \, y)$.
In particular, let
$$
\mathcal{Q}(1, k) = \{z_d \in \mathbb{R}: H_k(z_d) = 0\}
$$
be the set of nodes in one-dimension, exactly the zeros of the Hermite polynomials $H_k(z_d)$ for $k \in \mathbb{N}$, and
$$
\omega(z_d) = \frac{k!}{H_{k + 1}(z_d)^2 \times \phi(z_d)}, 
$$
be the associated weights.
Define $\mathcal{Q}(d, k) = \mathcal{Q}(1, k)^d$ to be the set of nodes $z$ in $d$-dimensions, of which there will be $k^d$.
Then these points are adapted via $\theta(z) = \hat \theta + Lz$, where $\hat \theta$ is the mode and $L$ is the lower Cholesky triangle of the curvature at the mode such that $LL^\top = H^{-1}$.

@@

Check here that `obj$fn` indeed takes as input $(x_i, \theta)$:

@@

transformed_param <- transform_param(param)
transformed_param$phi_prev_minus_i <- NULL #' Remove the integrated out effects

#' transformed_param contains certain values for (x_i, theta) suitable for input to obj$fn
obj$fn(transformed_param)
```

@@

ks_results %>%
  select(index, D, type) %>%
  pivot_wider(
    names_from = type,
    values_from = D
  ) %>%
  ggplot(aes(x = laplace, y = gaussian)) +
    geom_point(alpha = 0.5) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
    lims(x = c(0, 1), y = c(0, 1)) +
    labs(x = "KS(laplace, tmbstan)", y = "KS(gaussian, tmbstan)") +
    theme_minimal()

#' Table of average KS distance by method
ks_results %>%
  group_by(type) %>%
  summarise(Distance = mean(D)) %>%
  select(Distance) %>%
  mutate(
    Method = c("$\\text{KS}(\\texttt{tmbstan}, \\texttt{gaussian})$",
               "$\\text{KS}(\\texttt{tmbstan}, \\texttt{laplace})$"),
    .before = Distance
  ) %>%
  kableExtra::kable(booktabs = TRUE, escape = FALSE, align = "c")
```

@@

To start with, let's evaluate this on the log-scale for `i = 1`, `x = x_test_value` and the first value of `z`:

* The weight $\omega(z)$ will be `modesandhessians$weights[z]`
* The normalised Laplace approximation $\tilde p_\text{LA}(\theta(z) \, | \, y)$ will be `- h_norm(as.numeric(modesandhessians[z, thetanames]))`
* For the normalised Gaussian approximation $p_\text{G}(x_i \, | \, \theta(z), y)$ on the log-scale we get the relevant subcomponents of the joint Gaussian then use `ndorm` with `log = TRUE`

@@

lps <- sapply(gg$nodes, laplace_marginal)
plot_marginal_spline(lps)
```
## Full Laplace method (new new method)

Try doing this without a new `MakeADFun` for each $z$:

```{r}
new_laplace_marginal <- function(x, i) {
  dat_i <- prepare_dat(dat, i)

  obj <- MakeADFun(
    data = dat_i,
    parameters = param_i,
    random = "phi_prev_minus_i",
    DLL = "model1_index",
    silent = TRUE
  )

  lp <- vector(mode = "numeric", length = nrow(modesandhessians))

  for(z in 1:nrow(modesandhessians)) {
    theta <- as.numeric(modesandhessians[z, thetanames])
    lp[z] <- as.numeric(- obj$fn(c(theta[1], x, theta[2]))) + log(modesandhessians$weights[z])
  }

  matrixStats::logSumExp(lp) - quad$normalized_posterior$lognormconst
}
```

We get the same value as we did with `laplace_marginal`!

```{r}
new_laplace_marginal(x_test_value, i = 1)

lps <- sapply(gg$nodes, new_laplace_marginal, i = 1)
plot_marginal_spline(lps)
```
